Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-01-26 Response.tex

% 1c
\item {
{\it
\textbf{
Give a bit more detail on phylogeography, e.g. how do the models look like and what is the connection between phylogeography and epoch models (some connection is described in Chapter 2, but remains a bit vague).
}% END: textbf
}% END: it
}% END: item
\\
FB: I don't understand this comment, we have a whole subsection on phylogeography (see Subsection~\ref{sub:phylogeo}).

\end{itemize}

% 2
\item {
{\it
\textbf{
It is sometimes unclear in Chapter 2 whether the analysis is performed under a fixed evolutionary tree or not. 
This is especially so in the evaluation of Bayes Factors in Table 2.3. 
It would also have been nice to understand better whether the use of the epoch model (versus not) has an essential impact on tree reconstruction. 
Somewhat related, it would have been nice to learn from the simulation study how mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model. 
I am not asking for additional analyses here, but merely for additional clarification or insight.
}% END: textbf
}% END: it
}% END: item
\\
AAA

% 3
\item {
{\it
\textbf{
P6, lines 5-7: this is not correctly phrased as the parameter values do not have an associated probability in the frequentist likelihood framework.
}% END: textbf
}% END: it
}% END: item
\\
This is now rephrased:

\begin{quote}
\myeditsvthree
\end{quote}

% 4
\item {
{\it
\textbf{
Expression (1.12) is confusing in that indices i and j appear in the lefthand side, but not in the righthand side. 
Likewise in expression (1.13).
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 5
\item {
{\it
\textbf{
P20, 2nd display: an equality sign is missing on the 2nd line.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 6
\item {
{\it
\textbf{
I would find it useful to see a brief discussion on the plausibility of (1.22).
}% END: textbf
}% END: it
}% END: item
\\
AAA
% FB: he means from a biological perspective or statistical? It's a neccessary condition for the existence of stationary dist

% 7
\item {
{\it
\textbf{
If I am right that matrix P(t) on the bottom of page 22 is four-dimensional, then why not write it in full, which will be less confusing?
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 8
\item {
{\it
\textbf{
P25, the statement at the end of the penultimate paragraph (`most plausible hypothesis for the data') is vague.
}% END: textbf
}% END: it
}% END: item
\\
This statement now reads:

\begin{quote}
\myeditsveight
\end{quote}

% 9
\item {
{\it
\textbf{
P28, I understand what T(n) is, but would nevertheless find it helpful if it were stated more explicitly.
}% END: textbf
}% END: it
}% END: item
\\
AAA
%FB: I don't understand the question in the definition we clearly define T as function on a positive subset of real numbers.

% 10
\item {
{\it
\textbf{
P31, If the phrase `proper prior' is used, then perhaps it should be defined.
}% END: textbf
}% END: it
}% END: item
\\
This paragraph now reads:

\begin{quote}
\myeditsvnine
\end{quote}

% 11
\item {
{\it
\textbf{
P32, Since the author assumes that the reader may have no prior knowledge of Bayesian inference, I believe the meaning of Figure 1.11 should be explained in more detail.
}% END: textbf
}% END: it
}% END: item
\\
Corrected. There is now a more elaborate description.

% 12
\item {
{\it
\textbf{
P33,. I think the notation $\theta[t]$ has not been defined.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 13
\item {
{\it
\textbf{
There is a typo in the lefthand side of (1.37).
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 14
\item {
{\it
\textbf{
I would find it useful to get a bit more detail how a model for the likelihood of Y and for the prior of $\Phi$ might look like.
}% END: textbf
}% END: it
}% END: item
\\
AAA
%FB: We do write that the likelihood is standard tree pruning, we could elaborate on the prior choice here...

% 15
\item {
{\it
\textbf{
P43. What is compositional heterogeneity?
}% END: textbf
}% END: it
}% END: item
\\
We now exlipcitely state that we mean across-data (site) heterogeneity:

\begin{quote}
\myeditsvfifteen
\end{quote}

@@

% 16
\item {
{\it
\textbf{
P44. I would find it useful to receive a bit more guidance concerning the connection to phylogeographic inference. 
Also in Section 2.4.3, the description of the proposal by Bahl et al. (2011) is too brief for me to appreciate why the epoch model is more appropriate.
}% END: textbf
}% END: it
}% END: item
\\
AAA

% 17
\item {
{\it
\textbf{
Display (2.5): I believe this display implicitly assumes that the time T1 is a priori known. 
It would be good to state this explicitly.
}% END: textbf
}% END: it
}% END: item
\\
AAA
% FB: I don't understand the comment, I though we explicitely say we use spring and summer as intervals?

% 18
\item {
{\it
\textbf{
P57. The sentence `To determine the posterior odds, ... of that model' is vague.
}% END: textbf
}% END: it
}% END: item
\\
The sentence is rephrased:

\begin{quote}
\myeditsveighteen
\end{quote}

@@

% 19
\item {
{\it
\textbf{
P85. Riemannian -> Riemann
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 20
\item {
{\it
\textbf{
P86. The last sentence of Section 5.1.3 is not sufficiently clear. 
Please make clear why the use of mixed effects models could allow this.
}% END: textbf
}% END: it
}% END: item
\\
I changed the Subsection~\ref{sub:glm_epoch}, copied some bits from the hierarchical epoch model paper (under construction) and explained we mean th eimpact of various covariates as well as that by using MG94 parametrization we get a better resolution on selection parameters.

### 2015-01-26 chapters/chapter06.tex
discussing HMM

\subsection{Introducing covariates in the epoch model}

In order to test which factors explain the variation in an evolutionary parameter over time, we are currently exploring a \emph{mixed effects} modeling approach. 
Using hierarchical modeling we can introduce random effects for the evolutionary response variable \citep{Edo-Matas2011}.
Assuming effects are additive on the log-scale, we can complete the mixed effects model by incorporating fixed effects to model the impact of covariates on the evolutionary parameters. 
This represents an important data integration step, for which we have ample experience \citep{Edo-Matas2011,Streicker2012,Vrancken2014} and that will, for example, provide the opportunity to assess the association between within-host HIV evolutionary dynamics and clinical parameters (such as CD4 counts or viral load). 
This will lead to more direct testing of hypotheses that attempt to explain HIV divergence stabilization close to the onset of AIDS (e.g., the cellular exhaustion or immune relaxation hypotheses, \citep{Lemey2007}).

@@

%FB: typical notation  DP(G_0,\gamma)
Although the epoch model is useful as a specific case of allowing for variation in $\alpha$ and $\beta$ throughout evolutionary history, more general approaches would be desirable.
We are currently developing novel Bayesian non-parametric priors specifically tailored for evolutionary problems to hold the middle ground between the homogeneous MG94 model and estimating a separate MG94 model for each branch/site.
Bayesian non-parametric priors are often employed in problems where one wishes to cluster unknown parameters into distinct classes without making strong distributional assumptions.
The most common is the \emph{Dirichlet process} (DP) prior. 
In phylogenetics, this model has been developed for modeling among-site variation in the rates of non-synonymous substitutions \citep{Huelsenbeck2006} and to model lineage-specific substitution rates by assigning branches to rate categories \citep{Heath2012}.
A Dirichlet process DP$(F, \alpha)$, where $F$ (the base distribution) is a single parametric distribution and $\alpha$ (the concentration parameter) is a positive real number, returns a random distribution that is an infinite mixture of $F$ with different base parameters, e.g.~means and variances.
The partitioning of branches into specific realized values, called categories, is controlled by the concentration parameter $\alpha$, where small values of $\alpha$ lead to fewer categories and greater homogeneity of the substitution process, whereas large values indicate increased branch- or site-specific substitution processes.
Whereas often a single (branch- or site-specific) parameter is drawn from the base distribution $F$, our approach for the MG94 model requires two of such parameters to be drawn, for which we propose to first adapt the suggested DP prior.
A limitation of the DP is that there is no natural ordering for its realizations; any two draws share the same correlation.  
On an evolutionary history, this assumption is unrealistic; Thorne et al.~\citep{Thorne1998}, Seo et al.~\citep{SKT04} and Aris-Brosou and Yang \citep{Aris-Brosou2002} have long argued for phylogenetic auto-correlation among branch-specific rates.
We propose to achieve both auto-correlation and non-parametric flexibility by extending DPs to have first-order dependence along the evolutionary history.  
To accomplish this, recent work in Bayesian inference with DPs and their generalization, \emph{Polya urns}, \citep{Guha2010} shows that \emph{infinite hidden Markov} models \citep{Beal2002} in the computer science literature are now as convenient to fit as DPs and will allow for dependence across neighboring branches while retaining their non-parametric behavior.

>>>>

\subsection{Introducing covariates in the epoch model\label{sub:glm_epoch}}

In order to test which factors explain the variation in an evolutionary parameter over time, we are currently exploring a \emph{mixed effects} modeling approach. 
Using hierarchical modeling we can introduce random effects for the evolutionary response variable \citep{Edo-Matas2011}.
Assuming effects are additive on the log-scale, we can complete the mixed effects model by incorporating fixed effects to model the impact of covariates on the evolutionary parameters.

Specifically we parametrize each epoch specific, possibly multidimensional parameter $\Theta_{i},\; i=1,\ldots,M$ as a log-linear function of $P$ predictors $\mathbf{X}=\left(\mathbf{x_1},\ldots,\mathbf{x_P} \right)$:

\begin{align}
log\left(\Theta_{i}\right)=\beta \mathbf{X} \delta+\varepsilon_{i},\; i=1,\ldots,M,
\label{eq:mixed_model}
\end{align}

\noindent
where following the standard terminology in hierarchical modeling, the random effects $\varepsilon_{i}$ quantify different parameter values for each epoch.
$\mathbf{\beta}=\left(\beta_1,\ldots,\beta_P \right)^T$ represent the fixed effects and quantify contribution of a particular predictor to the dependent variable and we assume that:

\begin{equation}
\beta \sim \text{MVN}(\boldsymbol{\mu},\boldsymbol{\Sigma})
\label{eq:beta_proposal}
\end{equation}

\noindent
The $\left(\delta_1,\ldots,\delta_P \right)$ are binary indicator variables, on which we put independent Bernoulli priors:

\begin{equation}
\delta_{i} \sim \text{B}\left(1,p\right),\;i=1,\ldots,P.
\label{eq:indicator_prior}
\end{equation}

\noindent
The prior specification in Equation~\ref{eq:indicator_prior} allows for Bayesian stochastic search variable selection (BSSVS), which estimates posterior probabilities of inclusion or exclusion of a particular predictor in Equation~\ref{eq:glm} \cite{Lemey2009}.

This represents an important data integration step, for which we have ample experience \citep{Edo-Matas2011,Streicker2012,Vrancken2014} and that will, for example, provide the opportunity to assess the association between within-host HIV evolutionary dynamics and clinical parameters (such as CD4 counts or viral load). 
This will lead to more direct testing of hypotheses that attempt to explain HIV divergence stabilization close to the onset of AIDS (e.g., the cellular exhaustion or immune relaxation hypotheses, \citep{Lemey2007}) by allowing us to test the impact of various predictors in the course of the disease progression.
By using a different parametrization of codon model, such as the GY94 model, parametrized in terms of both the synonymous and non-synonymous rates of substitution, described in Subsection~\ref{sub:dpp} we would be able to disentangle the correlation between synonymous and non-synonymous selection parameters, something that offers an improvement over the approach presented in Chapter~\ref{chap:epoch}.

@@

%FB: typical notation  DP(G_0,\gamma)
Although the epoch model is useful as a specific case of allowing for variation in $\alpha$ and $\beta$ throughout evolutionary history, more general approaches would be desirable.
We are currently developing novel Bayesian non-parametric priors specifically tailored for evolutionary problems to hold the middle ground between the homogeneous MG94 model and estimating a separate MG94 model for each branch/site.
Bayesian non-parametric priors are often employed in problems where one wishes to cluster unknown parameters into distinct classes without making strong distributional assumptions.
The most common is the \emph{Dirichlet process} (DP) prior. 
In phylogenetics, this model has been developed for modeling among-site variation in the rates of non-synonymous substitutions \citep{Huelsenbeck2006} and to model lineage-specific substitution rates by assigning branches to rate categories \citep{Heath2012}.
A Dirichlet process DP$(F, \alpha)$, where $F$ (the base distribution) is a single parametric distribution and $\alpha$ (the concentration parameter) is a positive real number, returns a random distribution that is an infinite mixture of $F$ with different base parameters, e.g.~means and variances.
The partitioning of branches into specific realized values, called categories, is controlled by the concentration parameter $\alpha$, where small values of $\alpha$ lead to fewer categories and greater homogeneity of the substitution process, whereas large values indicate increased branch- or site-specific substitution processes.
Whereas often a single (branch- or site-specific) parameter is drawn from the base distribution $F$, our approach for the MG94 model requires two of such parameters to be drawn, for which we propose to first adapt the suggested DP prior.
A limitation of the DP is that there is no natural ordering for its realizations; any two draws share the same correlation.  
On an evolutionary history, this assumption is unrealistic; Thorne et al.~\citep{Thorne1998}, Seo et al.~\citep{SKT04} and Aris-Brosou and Yang \citep{Aris-Brosou2002} have long argued for phylogenetic auto-correlation among branch-specific rates.
We propose to achieve both auto-correlation and non-parametric flexibility by extending DPs to have first-order dependence along the evolutionary history.  
To accomplish this, recent work in Bayesian inference with DPs and their generalization, \emph{Polya urns}, \citep{Guha2010} shows that \emph{infinite hidden Markov models} \citep{Beal2002} in the computer science literature are now as convenient to fit as DPs and will allow for dependence across neighboring branches while retaining their non-parametric behavior.

%TODO: explain hidden markov models
\begin{remark}{Infinite hidden Markov models}
Hidden Markov model (HMM) defines a probability distribution over sequences of observations

\end{remark}
